---
title: "P"
author: "L"
date: "2025-07-18"
output: html_document
---

# 📄 Column Descriptions

-   **`id`**\
    Row ID consisting of the participant number and a count for that participant.

-   **`p_num`**\
    Participant number.

-   **`time`**\
    Time of day in the format `HH:MM:SS`.

-   **`bg-X:XX`**\
    Blood glucose reading in mmol/L, where `X:XX (H:MM)` is the time in the past\
    (e.g., `bg-2:35` means the reading from **2 hours and 35 minutes before** the time value for that row),\
    recorded by the continuous glucose monitor.

-   **`insulin-X:XX`**\
    Total insulin dose received in **units** in the last 5 minutes. `X:XX` refers to the time in the past\
    (e.g., `insulin-2:35` is the insulin dose **between 2:40 and 2:35 before** the time value),\
    recorded by the insulin pump.

-   **`carbs-X:XX`**\
    Total carbohydrate intake in **grams** in the last 5 minutes. `X:XX` indicates the time window in the past\
    (e.g., `carbs-2:35` is the amount **between 2:40 and 2:35 before** the time value),\
    recorded by the participant.

-   **`hr-X:XX`**\
    Mean heart rate in **beats per minute** in the last 5 minutes. `X:XX` represents the past time window\
    (e.g., `hr-2:35` is the mean heart rate **between 2:40 and 2:35 before** the time value),\
    recorded by the smartwatch.

-   **`steps-X:XX`**\
    Total steps walked in the last 5 minutes. `X:XX` indicates the time in the past\
    (e.g., `steps-2:35` = steps **between 2:40 and 2:35 before** the row's time),\
    recorded by the smartwatch.

-   **`cals-X:XX`**\
    Total calories burned in the last 5 minutes. `X:XX` refers to time in the past\
    (e.g., `cals-2:35` = calories **burned between 2:40 and 2:35 before** the row time),\
    calculated by the smartwatch.

-   **`activity-X:XX`**\
    Self-declared activity performed in the last 5 minutes. `X:XX` refers to time in the past\
    (e.g., `activity-2:35` shows the **activity name between 2:40 and 2:35 before** the row time),\
    set manually on the smartwatch.

-   **`bg+1:00`**\
    Blood glucose reading in mmol/L **one hour in the future** — this is the **target variable** to be predicted\
    (note: not provided in `test.csv`).

### Loading dataset

```{r}
library(dplyr)

sample <- read.csv("sample_submission.csv", header = TRUE) ### Sample
test <- read.csv("test.csv", header = TRUE) ### Test
train <- read.csv("train.csv", header = TRUE) ### Train

```

# understanding dataset

# check null

# check same values

# confirm number of people uniques by pnum

```{r}
dim(sample)
dim(test)
dim(train)

#str(train)
#glimpse(train)
#summary(train)
#install.packages("dplyr")
#install.packages("DT")

library(dplyr)
library(purrr)
library(DT)
# Assuming your data frame is named df
df<-train
total_rows <- nrow(df)

```

### merging dataset

```{r}

df_full <- bind_rows(train, test)
dim(df_full)

```
### reduce 72 columns per each relevan category to avgs
```{r}
library(readr)
# Columns by prefix
bg_cols <- grep("^bg.", names(train), value = TRUE)
insulin_cols <- grep("^insulin.", names(train), value = TRUE)
cat("bg columns:", length(bg_cols), "\n")
cat("insulin columns:", length(insulin_cols), "\n")

# Variables to average in blocks
variables <- c("bg", "carbs", "hr", "steps", "cals")

# Compute block averages (24 blocks of 3 columns)
for (var in variables) {
  var_cols <- grep(paste0("^", var, "."), names(train), value = TRUE)
  
  for (i in 0:23) {
    cols_3 <- var_cols[(i*3 + 1):(i*3 + 3)]
    avg_col <- paste0(var, "_avg_", i)
    train[[avg_col]] <- rowMeans(train[, cols_3], na.rm = TRUE)
  }
}

# Check avg columns
avg_cols <- grep("_avg_", names(train), value = TRUE)

# Create cleaned dataset
df_clean <- train %>%
  select(p_num, time, all_of(avg_cols)) %>%
  mutate(target = train[["bg+1:00"]])

# Check missing in target
cat("Missing target values:", sum(is.na(df_clean$target)), "\n")

# Save clean data
write_csv(df_clean, "data/df_clean.csv")

# Load and structure cleaned dataset
df <- read_csv("data/df_clean.csv")
glimpse(df)

# Convert time to POSIXct and extract hour/minute
df <- df %>%
  mutate(
    time = hms::as_hms(time),
    hour = hour(time),
    minute = minute(time)
  )

# Create period of day
df <- df %>%
  mutate(
    period = case_when(
      hour < 6 ~ "night",
      hour < 12 ~ "morning",
      hour < 18 ~ "afternoon",
      TRUE ~ "evening"
    )
  )

# Create day_id per p_num
df <- df %>%
  mutate(
    time_seconds = hour * 3600 + minute * 60 + second(time)
  ) %>%
  group_by(p_num) %>%
  arrange(p_num, time_seconds) %>%
  mutate(
    time_diff = time_seconds - lag(time_seconds, default = first(time_seconds)),
    new_day = if_else(time_diff < 0, 1, 0),
    day_id = cumsum(new_day)
  ) %>%
  ungroup()

# View result
print(df)

# Save final dataset
write_csv(df, "data/df_check.csv")

```

```{r}

dim(df_full2)

```

