---
title: "CLARA Example"
author: "DANA 4840"
output:
  pdf_document: default
  html_document: default
---

### Generating data for example

1. Split randomly the data sets in multiple subsets with fixed size (sampsize)

```{r}
set.seed(1234)
# Generate 500 objects, divided into 2 clusters.
df <- rbind(cbind(rnorm(200,0,8), rnorm(200,0,8)),
           cbind(rnorm(300,50,8), rnorm(300,50,8)))

```


```{r}
# Specify column and row names
colnames(df) <- c("x", "y")
rownames(df) <- paste0("S", 1:nrow(df))
```


```{r}
# Previewing the data
head(df, nrow = 6)
```

```{r}
library(cluster)
library(factoextra)
```



```{r}

library(cluster)
library(factoextra)
```

### Estimating the optimal number of clusters

```{r}
fviz_nbclust(df, clara, method = "silhouette")+
  theme_classic()
```



From the plot, the suggested number of clusters is 2. In the next section, we'll classify the observations into 2 clusters.


### Computing CLARA


```{r}
clara.res <- clara(df, 2, samples = 50, pamLike = TRUE)
```



```{r}
print(clara.res)
```

If you want to add the point classifications to the original data, use this:

```{r}
dd <- cbind(df, cluster = clara.res$cluster)
head(dd, n = 4)
```


You can access to the results returned by clara() as follow:

```{r}
# Medoids
clara.res$medoids
```


```{r}
# Clustering
head(clara.res$clustering, 10)
```



### Visualizing CLARA clusters

```{r}
fviz_cluster(clara.res, 
             palette = c("#00AFBB", "#FC4E07"), # color palette
             ellipse.type = "t", # Concentration ellipse
             geom = "point", pointsize = 1,
             ggtheme = theme_classic()
             )

```



