---
title: "Cluster distances"
author: "Example 1"
output:
  pdf_document: default
  html_document: default
---
We'll use the USArrests data as demo data sets. We'll use only a subset of the data
by taking 15 random rows among the 50 rows in the data set. This is done by using
the function _sample()_. Next, we standardize the data using the function _scale()_:

#####Scaling data
```{r}
#Subset of the data
set.seed(123)
ss <- sample(1:50, 15) # Take 15 random rows
df <- USArrests[ss, ] # Subset the 15 rows
df.scaled <- scale(df) # Standardize the variables
```

###Computing euclidean distance

We use the _dist()_ function:

```{r}
library(stats)
dist.eucl <- dist(df.scaled, method = "euclidean")
#dist.eucl
```

#####What other methods can be used for the _dist()_ function?

To make it easier to see the distance information generated by the _dist()_ function, you can reformat the distance vector into a matrix using the _as.matrix()_ function.

```{r }
# Reformat as a matrix
# Subset the first 3 columns and rows and Round the values
round(as.matrix(dist.eucl)[1:3, 1:3], 1)
```

#####What do these values represent?

In this data set, the columns are variables. Hence, if we want to compute pairwise distances between variables, we must start by transposing the data to have variables in the rows of the data set before using the _dist()_ function. The function _t()_ is used for transposing the data.


###Computing correlation based distances

```{r}
# install.packages(factoextra)
library("factoextra")
dist.cor <- get_dist(df.scaled, method = "pearson")
```

```{r}
# Display a subset
round(as.matrix(dist.cor)[1:3, 1:3], 1)
```
#####What other methods can be used for the _dist()_ function?

###Computing distances for mixed data

The function _daisy()_ [ _cluster_ package] provides a solution (Gower's metric) for computing the distance matrix, in the situation where the data contain no-numeric columns.


The R code below applies the _daisy()_ function on _flower_ data which contains factor, _ordered_ and _numeric_ variables:

```{r}
#install.packages(cluster)
library(cluster)

# Load data
data(flower)
head(flower, 3)
```


```{r}
# Data structure
str(flower)
```


```{r}
# Distance matrix
dd <- daisy(flower)
round(as.matrix(dd)[1:3, 1:3], 2)
```



###Visualizing distance matrices

To use _fviz_dist()_ type this:

```{r}
library(factoextra)
fviz_dist(dist.eucl)
```


Red: high similarity (ie: low dissimilarity) | Blue: low similarity.
