---
title: "Cluster Algorithms"
author: "DANA 4840"
output:
  word_document: default
  html_document: default
---


```{r}
#clValid(obj, nClust, clMethods = "hierarchical",
      #   validation = "stability", maxitems = 600,
      #  metric = "euclidean", method = "average")
```

- **obj**: A numeric matrix or data frame. Rows are the items to be clustered and columns are samples.

```{r, message=FALSE}
library(clValid)
# Iris data set:
# - Remove Species column and scale
df <- scale(iris[, -5])
```


#### Internal Measures
```{r}
# Compute clValid
clmethods <- c("hierarchical","kmeans","pam")

intern <- clValid(df, nClust = 2:6, 
              clMethods = clmethods, validation = "internal")
# Summary
summary(intern)
```



It can be seen that hierarchical clustering with two clusters performs the best in each case (i.e., for connectivity, Dunn and Silhouette measures). Regardless of the clustering algorithm, the optimal number of clusters seems to be two using the three measures.

#### Stability Measures
```{r}
# Stability measures
clmethods <- c("hierarchical","kmeans","pam")
stab <- clValid(df, nClust = 2:6, clMethods = clmethods, 
                validation = "stability")
# Display only optimal Scores
optimalScores(stab)
```


For the APN and ADM measures, hierarchical clustering with two clusters again gives the best score. For the other measures, PAM with six clusters has the best score.

