---
title: "Comparing and Visualing Dendrogram"
author: "DANA 4840"
date: "27/09/2022"
output:
  pdf_document: default
  html_document: default
---
## Prepare Data

```{r}
df <- scale(USArrests)
```

```{r}
# Subset containing 10 rows
set.seed(123)
ss <- sample(1:50, 10)
df <- df[ss,]
```


#### Comparing Dendrograms

```{r}
library(dendextend)

# Compute distance matrix
res.dist <- dist(df, method = "euclidean")

# Compute 2 hierarchical clusterings using 2 methods
hc1 <- hclust(res.dist, method = "average")
hc2 <- hclust(res.dist, method = "ward.D2")

# Create two dendrograms
dend1 <- as.dendrogram (hc1)
dend2 <- as.dendrogram (hc2)

# Create a list to hold dendrograms
dend_list <- dendlist(dend1, dend2)
```



### Visual Comparison of two dendrograms

```{r}
tanglegram(dend1, dend2)
```



```{r}

#Customized the tanglegram using many other options as follow:
tanglegram(dend1, dend2,
highlight_distinct_edges = FALSE, # Turn-off dashed lines
common_subtrees_color_lines = FALSE, # Turn-off line colors
common_subtrees_color_branches = TRUE, # Color common branches
main = paste("entanglement =", round(entanglement(dend_list), 2))
)                   # Alignment quality
```

Note that, "unique" nodes, with a combination of labels/items not present in the other tree, are highlighted with dashed lines.

#### Correlation matrix between a list of dendrograms

The function cor.dendlist() is used to compute “Baker” or “Cophenetic” correlation matrix between a list of trees. The value can range between -1 to 1. With near 0 values meaning that the two trees are not statistically similar.


```{r}
# Cophenetic correlation matrix
cor.dendlist(dend_list, method = "cophenetic")
```


```{r}
# Baker correlation matrix
cor.dendlist(dend_list, method = "baker")
```



```{r}
# Cophenetic correlation coefficient
cor_cophenetic(dend1, dend2)
```


```{r}
# Baker correlation coefficient
cor_bakers_gamma(dend1, dend2)
```

It’s also possible to compare simultaneously multiple dendrograms. A chaining operator
%>% is used to run multiple function at the same time. It’s useful for simplifying the
code:

```{r}
# Create multiple dendrograms by chaining
dend1 <- df %>% dist %>% hclust("complete") %>% as.dendrogram
dend2 <- df %>% dist %>% hclust("single") %>% as.dendrogram
dend3 <- df %>% dist %>% hclust("average") %>% as.dendrogram
dend4 <- df %>% dist %>% hclust("centroid") %>% as.dendrogram
```


```{r}
# Compute correlation matrix
dend_list <- dendlist("Complete" = dend1, "Single" = dend2,
                      "Average" = dend3, "Centroid" = dend4)

```


```{r}
cors <- cor.dendlist(dend_list)
# Print correlation matrix
round(cors, 2)
```


```{r}
# Visualize the correlation matrix using corrplot package
library(corrplot)
corrplot(cors, "pie", "lower")
```


## Visualizing Dendrograms


```{r}
# Load data
data(USArrests)
```


```{r}
# Compute distances and hierarchical clustering
dd <- dist(scale(USArrests), method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
```

```{r}
## install.packages(c("factoextra", "dendextend"))
library(factoextra)
fviz_dend(hc, cex = 0.5)

```

You can use the arguments main, sub, xlab, ylab to change plot titles as follow:
```{r}
fviz_dend(hc, cex = 0.5,
           main = "Dendrogram - ward.D2",
          xlab = "Objects", ylab = "Distance", sub = "")
```

To draw a horizontal dendrogram, type this:
```{r}
fviz_dend(hc, cex = 0.5, horiz = TRUE)
```


It’s also possible to cut the tree at a given height for partitioning the data into multiple groups as described in the previous chapter: Hierarchical clustering. In this case, it’s possible to color branches by groups and to add rectangle around each group.

For example:

```{r}
fviz_dend(hc, k = 4, # Cut in four groups
          cex = 0.5, # label size
          k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
          color_labels_by_k = TRUE, # color labels by groups
          rect = TRUE, # Add rectangle around groups
          rect_border = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"), 
          rect_fill = TRUE)
```


```{r}
fviz_dend(hc, k = 4,                 # Cut in four groups
         cex = 0.5,                 # label size
          k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
          color_labels_by_k = TRUE,  # color labels by groups
         ggtheme = theme_gray()     # Change theme
          )
```


```{r}
fviz_dend(hc, cex = 0.5, k = 4, # Cut in four groups
          k_colors = "jco")
```



```{r}

fviz_dend(hc, k = 4, cex = 0.4, horiz = TRUE,  k_colors = "jco", 
          rect = TRUE, rect_border = "jco", rect_fill = TRUE)

```


```{r}

fviz_dend(hc, cex = 0.5, k = 4, 
          k_colors = "jco", type = "circular")
```

To plot a phylogenic-like tree, use type = “phylogenic” and repel = TRUE (to avoid
labels overplotting). This functionality requires the R package *igraph*.


```{r}
require("igraph")
fviz_dend(hc, k = 4, k_colors = "jco",
          type = "phylogenic", repel = TRUE)
```



```{r}
require("igraph")
fviz_dend(hc, k = 4, # Cut in four groups
          k_colors = "jco",
          type = "phylogenic", repel = TRUE,
          phylo_layout = "layout.gem")
```


## Zooming in the dendrogram
```{r}
fviz_dend(hc, xlim = c(1, 20), ylim = c(1, 8))
```


### Plotting a sub-tree of dendrograms


```{r}
# Create a plot of the whole dendrogram,
# and extract the dendrogram data
dend_plot <- fviz_dend(hc, k = 4, # Cut in four groups
          cex = 0.5, # label size
          k_colors = "jco"
          )
dend_data <- attr(dend_plot, "dendrogram") # Extract dendrogram data

```


```{r}
# Cut the dendrogram at height h = 10
dend_cuts <- cut(dend_data, h = 10)
# Visualize the truncated version containing
# two branches
fviz_dend(dend_cuts$upper)
```


```{r}
# Plot the whole dendrogram
print(dend_plot)
```


```{r}
# Plot subtree 1
fviz_dend(dend_cuts$lower[[1]], main = "Subtree 1")
```



```{r}
# Plot subtree 2
fviz_dend(dend_cuts$lower[[2]], main = "Subtree 2")
```


```{r}
fviz_dend(dend_cuts$lower[[2]], type = "circular")
```




## Saving dendrogram into a large PDF page
```{r}
## pdf("dendrogram.pdf", width=30, height=15)            # Open a PDF
## p <- fviz_dend(hc, k = 4, cex = 1, k_colors = "jco" ) # Do plotting
## print(p)
## dev.off()                                             # Close the PDF
```






