---
title: "DBSCAN"
author: "DANA 4840"
output:
  word_document: default
  html_document: default
---

```{r}
# Load the data set
library("factoextra")
data("multishapes")
df <- multishapes[, 1:2]
# Visualize the data
ggplot(df, aes(x, y)) + geom_point()
```
##### Checking K-means
```{r}
set.seed(123)
km.res <- kmeans(df, 5, nstart = 25)
fviz_cluster(km.res, df, geom = "point",
ellipse= FALSE, show.clust.cent = FALSE,
palette = "jco", ggtheme = theme_classic())
```

```{r}
## Compute DBSCAN using fpc package
library("fpc")
set.seed(123)
db <- fpc::dbscan(df, eps = 0.15, MinPts = 5)
```



```{r}
# Plot DBSCAN results
library("factoextra")
fviz_cluster(db, data = df, stand = FALSE,
             ellipse = FALSE, show.clust.cent = FALSE,
             geom = "point",palette = "jco", ggtheme = theme_classic())
```

Note that, the function *fviz_cluster*() uses different point symbols for core points (i.e, seed points) and border points. Black points correspond to outliers. You can play with *eps* and *MinPts* for changing cluster configurations.

It can be seen that DBSCAN performs better for these data sets and can identify the correct set of clusters compared to k-means algorithms.
 

```{r}
print(db)
```
In the table above, column names are cluster number. Cluster 0 corresponds to outliers (black points in the DBSCAN plot). The function print.dbscan() shows a statistic of the number of points belonging to the clusters that are seeds and border points.

```{r}
db$cluster[sample(1:1089, 20)]

```

DBSCAN algorithm requires users to specify the optimal eps values and the parameter MinPts. In the R code above, we used eps = 0.15 and MinPts = 5. One limitation of DBSCAN is that it is sensitive to the choice of $\epsilon$, in particular if clusters have different densities. If $\epsilon$ is too small, sparser clusters will be defined as noise. If $\epsilon$ is too large, denser clusters may be merged together. This implies that, if there are clusters with different local densities, then a single $\epsilon$ value may not suffice.

A natural question is:

 How to define the optimal value of \textit{eps}?


The function kNNdistplot() [in dbscan package] can be used to draw the k-distance plot:

```{r}
dbscan::kNNdistplot(df, k =  5)
abline(h = 0.15, lty = 2)
```

