---
title: "K-means"
author: "Example"
output:
  pdf_document: default
  html_document: default
---
In this article, we'll describe the **k-means algorithm** and provide practical examples using **R** software. We will use the demo data sets "USArrests".

####1. Specify the number of clusters (K) to be created (by the analyst)

```{r}
data("USArrests")      # Loading the data set
df <- scale(USArrests) # Scaling the data
```

#### View the firt 3 rows of the data
```{r}
head(df, n = 3)
```


####K-means function in R and arguments

kmeans(x, centers, iter.max = 10, nstart = 1)

- **x**: numeric matrix, numeric data frame or a numeric vector

```{r}
## install.packages("factoextra")

library(factoextra)
```


####Estimating the optimal number of clusters
The k-means clustering requires the users to specify the number of clusters to be generated.

One fundamental question is: How to choose the right number of expected clusters (k)?


Different methods will be presented in the chapter "cluster evaluation and validation statistics".


Here, we provide a simple solution. The idea is to compute k-means clustering using
different values of clusters k. Next, the wss (within sum of square) is drawn according to the number of clusters. The location of a bend (knee) in the plot is generally considered as an indicator of the appropriate number of clusters.

The R function _fviz_nbclust()_ [in _factoextra_ package] provides a convenient solution to estimate the optimal number of clusters.

```{r}
fviz_nbclust(df, kmeans, method = "wss") +
    geom_vline(xintercept = 4, linetype = 2)
```


The plot above represents the variance within the clusters. It decreases as k increases, but it can be seen a bend (or "elbow") at k = 4. This bend indicates that additional clusters beyond the fourth have little value.. In the next section, we'll classify the observations into 4 clusters.


#### Compute k-means clustering with k = 4

```{r}
set.seed(123)
km.res <- kmeans(df, 4, nstart = 25)
```

As the final result of k-means clustering result is sensitive to the random starting assignments, we specify *nstart = 25*. This means that R will try 25 different random starting assignments and then select the best results corresponding to the one with the lowest within cluster variation. The default value of *nstart* in R is one. But, it's strongly recommended to compute *k-means clustering* with a large value of *nstart* such as 25 or 50, in order to have a more stable result.


#####Print the results
```{r}
print(km.res)
```

What do we see in the output?


It's possible to compute the mean of each variables by clusters using the original data:

```{r}
aggregate(USArrests, by=list(cluster=km.res$cluster), mean)
```


If you want to add the point classifications to the original data, use this:

```{r}
dd <- cbind(USArrests, cluster = km.res$cluster)
head(dd)
```

####Accessing to the results of kmeans() function

#####Cluster number for each of the observations

```{r}
#km.res$cluster

head(km.res$cluster, 4)

```

#####Cluster size

```{r}
km.res$size
```

#####Cluster means

```{r}
km.res$centers
```

####Visualizing k-means clusters

If we have a multi-dimensional data set, a solution is to perform Principal Component Analysis (PCA) and to plot data points according to the first two principal components coordinates.


```{r}
fviz_cluster(km.res, data = df,
             palette = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"), 
             ellipse.type = "euclid", # Concentration ellipse
             star.plot = TRUE, # Add segments from centroids to items
             repel = TRUE, # Avoid label overplotting (slow)
             ggtheme = theme_minimal()
             )
```


