---
title: "Activity1"
author: "Maria Sandate"
date: "2025-01-14"
output:
  word_document: default
  html_document: default
---

#### Activity 1

2.	The weight of a specific product follows a normal distribution with a mean of 30 g and a standard deviation of 4 g.

- Calculate the probability that a randomly chosen product weights less than 27.6g.

```{r}

pnorm(27.6,30,4, lower.tail = TRUE )

```
 
- What is the probability that a randomly chosen product has a weight that exceeds 26.6 g? 

```{r}
pnorm(26.6,30,4, lower.tail = FALSE )

```

c. What is the probability that a randomly chosen product is between 26.9 and 29.6 g?

```{r}
a <- pnorm(26.9,30,4, lower.tail = TRUE )
b <- pnorm(29.6,30,4, lower.tail = TRUE )
b-a
```
3a. What is the maximum weight of the lightest 10% of products? 

```{r}
qnorm(0.1,30, 4, lower.tail = TRUE)
```

3b. What is the minimum weight of the heaviest 5% of products? 

```{r}
qnorm(0.05,30, 4, lower.tail = FALSE)
```
4a. calculates the z-score for the 10th percentile, which corresponds to the maximum 
weight of the lightest 10% of products. 

```{r}
z_score <- qnorm(0.1)
z_score
```

4b. calculates the z-score for the 95th percentile, which corresponds to the minimum 
weight of the heaviest 5% of products. 

```{r}
z_score <- qnorm(0.95)
z_score
```

5. We are interested to calculate the marginal error when we are 95% Confident that 
our interval contains the true mean of our population. What should be the z-score? 

```{r}
z_score <- qnorm(0.975)
z_score
```
6. The weight of a specific product follows a normal distribution with a mean of 30 g 
and a standard deviation of 4 g. 

6a. If we select a simple random sample of 49 products, what is the probability that 
the average weight of our sample is less than 30.6 g? 

```{r}

SE<-4/sqrt(49)

pnorm(30.6,30,SE, lower.tail = TRUE )
```

6b. If we select a simple random sample of 49 products, what is the probability that 
the average weight of our sample is at least 30.6 g? 

```{r}
pnorm(30.6,30,SE, lower.tail = FALSE )
```

6c. If we select a simple random sample of 49 products, what is the probability that 
the average weight of our sample is between 28.9 and 29.6 g 

```{r}
a <- pnorm(28.9,30,SE, lower.tail = TRUE )
b <- pnorm(29.6,30,SE, lower.tail = TRUE )
b-a
```

6d. Ninety percent of the average weight of 49 randomly selected products are less 
than how many grams?   

```{r}
qnorm(.9,30, SE, lower.tail = TRUE)
```


6e. Ninety percent of the average weight of 49 randomly selected product are greater 
than how many grams?

```{r}
qnorm(.1,30, SE, lower.tail = TRUE)
```

7. Psychologists have found that twins, in their early years, tend to have lower 
intelligence quotients and pick up language more slowly than nontwins. The slower 
intellectual growth of twins may be caused by benign parental neglect.  

Suppose we want to investigate this phenomenon. A random sample of n = 50 sets of 
2.5 -year-old twin boys is selected, and the total parental attention time given to each 
pair during 1 week is recorded. 

Estimate μ, the mean attention time given to all 2.5 year-old twin boys by their 
parents, using a 99% confidence interval. Interpret the interval in terms of the 
problem. (Use ATTENTIMES -> chapter1 -> D2L.) 

```{r}
ATTENTIMES <- read.csv("ATTENTIMES.csv", header=TRUE)
```

```{r}
head(ATTENTIMES)
```
```{r}
install.packages("BSDA")
```
```{r}
library(BSDA)

```
```{r}
View(ATTENTIMES)
attach(ATTENTIMES)
head(ATTENTIMES)
mean(Attention.Time)
```


8. The Geothermal Loop Experimental Facility, located in the Salton Sea in southern 
California, is a U.S. Department of Energy operation for studying the feasibility of 
generating electricity from the hot, highly saline water of the Salton Sea. Operating 
experience has shown that these brines leave silica scale deposits on metallic plant 
piping, causing excessive plant outages. Researchers found that scaling can be reduced 
somewhat by adding chemical solutions to the brine. In one screening experiment, each 
of five antiscalants was added to an aliquot of brine, and the solutions were filtered. A 
silica determination (parts per million of 
silicon dioxide) was made on each filtered sample after a holding time of 24 hours, 
with the results stored in SILICA data set (D2L).  Estimate the mean amount of 
silicon dioxide present in the five antiscalant solutions. Use a 95% confidence interval. 


9. After installing a drive-thru ATM on the outside of a local bank, the bank manager
wanted to see if customers’ waiting time inside the bank was shortened during lunch 
hour (12noon-1pm).  In the past (before the installation of the drive-thru ATM), the 
average customers’ waiting time (inside the bank) during lunch hour has been 15 
minutes.  From a random sample of 20 bank customers during lunch hour (12noon-1pm), 
the average wait time was found to be 12.3 minutes.  It is known that the wait time has a Normal distribution and a standard deviation of 9 minutes.  Can we show that the current wait time has been shortened? 
a. Identify the parameter of interest. 

The mean of the wating time of all customers

b. Set up the null hypothesis and the alternative hypothesis. 

Ho: mu >= 15
Ha: mu <15

P(Xbar < 12.3)

c. Draw an appropriate conclusion.  Use alpha= 0.05.


d. What if you wanted to check that waiting time ≠ 15? 


10. Humerus bones from the same species of animal tend to have approximately 
the same length-to-width ratios. When fossils of humerus bones are discovered, 
archeologists can often determine the species of animal by examining the lengthto- 
width ratios of the bones. It is known that species A has a mean ratio of 8.5. 
Suppose 41 fossils of humerus bones were unearthed at an archeological site in East 
Africa, where species A is believed to have flourished. (Assume that the unearthed 
bones were all from the same unknown species.) The length-to-width ratios of 
the bones were measured and are recorded in BONES data set. 

a. Do these data present sufficient evidence to indicate that the mean ratio of all 
bones of this species differs from 8.5? Use α = .05. 
b. The mean ratio of all bones of this species is greater than 8.5? Use α = .05. 
c. The mean ratio of all bones of this species is less than 8.5? Use α = .05. 
11. A dietitian has developed a diet that is low in fats, carbohydrates, and cholesterol. 
Although the diet was initially intended to be used by people with heart disease, 
the dietitian wishes to examine the effect this diet has on the weights of obese 
people. Two random samples of 100 obese people each are selected, and one group 
of 100 is placed on the low-fat diet. The other 100 are placed on a diet that contains 
approximately the same quantity of food but is not as low in fats, carbohydrates, 
and cholesterol. For each person, the amount of weight lost (or gained) in a 3-week 
period is recorded. The data, saved in DIETS data set (D2L). 
a. Form a 95% confidence interval for the difference between the population mean 
weight losses for the two diets. Interpret the result. 
b. What is the alternative hypothesis in a? 
c. What if you want to test that the differences is greater than or smaller than zero? 
12. Suppose you wish to compare a new method of teaching reading to ‘‘slow learners’’ 
to the current standard method. You decide to base this comparison on the results 
of a reading test given at the end of a learning period of 6 months. Of a random 
sample of 22 slow learners, 10 are taught by the new method and 12 are taught 
by the standard method. All 22 children are taught by qualified instructors under 
similar conditions for a 6-month period. The results of the reading test at the end of 
this period are recorded in READING data set.  
a. Use the data to test whether the true mean test scores differ for the new method 
and the standard method. Use α = .05. 
b. What assumptions must be made in order that the estimate be valid? 
13. Data set: PAIRED shows the data on reading test scores for eight pairs of slow 
learners with similar reading IQs of which one member is randomly assigned to the 
standard teaching method while the other is assigned to the new method. 
Do the data support the hypothesis that the population mean reading test score for 
slow learners taught by the new method is greater than the mean reading test score 
for those taught by the standard method.