---
title: "Mid"
author: "Maria Sandate"
date: "2025-02-14"
output:
  word_document: default
  html_document: default
---


#Data

```{r}
low.btw <- load("low_bwt.Rdata")

```



### Model completed

#a)
```{r}
model1 <- lm(birthwt~toxemia,)
summary(model1)
```

# i) Equation  birth weight = 1097.215 +7.785*Toxemia Yes

# ii) Confidence interval

```{r}
confint(model1,level=0.95)
```
### Interpretation

Since the confidence interval at 95% goes from a negative value (-124.42) and a positive one (139.9899) we can not be confident about the effect of the toxemia in the birth weight.

# b)

```{r}
model2 <- lm(birthwt~toxemia+gestage,)
summary(model2)
```

# i)

B0=-1286.2 is the value base when Toxemia is No ans it is meaningful because it our base point when we keep constant the gestational age.
B1=-206.591 when Toxemia is Yes, decrease the birth weight 206.591 grams when we keep gestational age constant.
B2=84.058 when the gestational age increase one week and the Toxemia is NO, the birth weight increase 84.058 grams.

# ii)

Equation 

birth weight = -1286.2 -206.591*(Toxemia Yes)+84.058*gestational age

```{r}
New_data=data.frame(toxemia=c("Yes"), gestage=c(31))
predict(model2, newdata = New_data, interval="prediction", level=0.95)
```
The average birth weight for an infant born to a mother diagnosed with toxemia with 31 weeks is predicted to have a weight of 1113 grams.

# ii)

```{r}
predict(model2, newdata = New_data, interval="confidence", level=0.95)
predict(model2, newdata = New_data, interval="prediction", level=0.95)
```

We estimate with a confidence of 95% that the mean value of E(y) is between 1030 an 1195.
We predict y_hat is between 727.9 an 1498.