---
title: "Mid"
author: "Leo L sanchez"
date: "2025-02-14"
output:
  word_document: default
  html_document: default
---


#Data

```{r}
# 1) Cargar el .Rdata (NO lo asignes)
load("low_bwt.Rdata")

# 2) Ver qué objeto cargó
ls()

# 3) (Asumiendo que el objeto se llama low_bwt)
head(low.bwt)
str(low.bwt)

# 4) Ajustar el modelo correctamente
model <- lm(birthwt ~ toxemia, data = low.bwt)

# 5) Resultado
summary(model)
```

write the model equation 

predicted birth weight = 1097.215 + 7.785toxemiaYes 

report a  95% confidence interval for the slope and interpret the interval
```{r}
  confint(model,level = 0.95)
```

we are 95% confidence interval for the slode is 1124.42,139.98
captures the population difference in mean between infants from mother diagnosted by toxemia   and infant with mothers no diagnosted by toxemia



```{r}
# 1) Cargar el .Rdata (NO lo asignes)
load("low_bwt.Rdata")

# 2) Ver qué objeto cargó
ls()

# 3) (Asumiendo que el objeto se llama low_bwt)
head(low.bwt)
str(low.bwt)

# 4) Ajustar el modelo correctamente
modelb <- lm(birthwt ~ toxemia+gestage , data = low.bwt)

# 5) Resultado
summary(modelb)
```

b0 -1286.200 
represent the value baseline when toxemia is no and our gestage also is 0  

b1 toxemiayes  206.591
represent  the effect of categorical variable toxemia, since "NO" is the baseline (absence variable) the coefficien toxemiaYes indicates the difference in the mean response between the two groups when we keep gestage constant

b2 gestage 84.058
holding toxemia variable fic¿xed, for every increase of gestage, the mean birht weight  is estimeate to increase by 84058

intercep is meaningful?
the intercep dooes not  have a practical interpretation since -1286 is not a real value for birthweight



